The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-08 pkg/partykit/inst/guideliketest/draft/draft.Rnw
corrected minor typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2547 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 19dd5ea266a9117a17589d924a4a468c7f6e5a84

\subsection{Generic algorithm}
\label{sec:generic_alg}
The basic idea of building a regression tree model is to partition the data into smaller 
and more homogeneous subgroups based on a set of covariates. Various tree 
algorithms have been developed, following essentially the same general structure.
Starting at the root of the tree, pertaining to the full available data sample, the
algorithms proceed in the following steps:
%
\begin{enumerate}
  \item For the current sample a (possibly simple) model is fitted by optimizing
        some objective function (or loss function) that reflects the goodness of fit.
  \item Among all available covariates one is selected as the split variable,
        choosing the split point such that goodness of fit is maximized in the
        resulting subgroups.
  \item Steps 1 and 2 are repeated within each subgroup until some stopping
        criterion is attained.
\end{enumerate}
%
The term ``model'' is used here in a very broad sense and encompasses not
only least-squares or maximum-likelihood models but also simple constant
fits such as means or average proportions.
Therefore, depending on the type of model, the employed objective function can 
for example be the sum of deviations from a typical/average value, but it can 
also be based on a model fro a response along with potential covariates.
For instance, different types of residuals (or the signs thereof) can be employd
\citep[see e.g.,][]{Loh:2002} as well as rank sums or logrank scores
\citep[as in][]{Hothorn+Hornik+VanDeWiel:2006}.

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and each possible transformed split variable $g(Z_j)$ ($j = 1, \dots, m$).
The only requirement for the function $h$ is to depend on $Y$ in a permutation-symmetric way but
this encompasses ranks, scores, indicator functions, etc.\ and can also be multidimensional.
For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can capture location and scale of $Y$ via
$h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fitted to the response $Y$ with regressor variable(s)~$X$,
then a model-based transformation $h(Y) = s(Y,X,\hat{\beta})$ can be used, e.g.,
the residuals in a linear model with regression coefficients $\beta$. Moreover,
$s$ can be the score function pertaining to the objective function (or loss function)
$\ell$:
$$
s(Y, X, \beta) = \frac{\partial \ell(Y, X, \beta)}{\partial \beta}
$$
The estimate of the model parameters $\hat{\beta}$ is obtained by optimizing
the sum of the objective function $\ell$, aggregated over all observations.
This framework includes many different M-type estimators as special cases, including maximum 
likelihood and ordinary least squares estimation. For a $k$-dimensional parameter
$\beta$ the score function evaluated for the $i$-th observation $s(y_i, x_i, \beta)$
is also a $k$dimensional vector, i.e., the gradient contribution of the $i$-th observation.
Thus, the $n \times k$-matrix consisting of these scores or gradient contributions for all 
$i = 1, \dots, n$ observations is a natural candidate for the observation-wise goodness-of-fit
measure as described in the previous Section~\ref{sec:generic_alg}.

@@

To test for independence of $h(Y)$ and $g(Z)$ CTree calculates a linear association test statistic,
following the framework of \cite{Strasser+Weber:1999}. The corresponding conditional expectation and
covariance given all permutations of the response variable can be calculated and used to 
standardize the test statistic. This standardized statistic has an asymptotic normal distribution
which is in fact multivariate if either of the transformation $h(Y)$ and/or $g(Z)$ is
multivariate. The actual test is carried out be mapping this standardized statistic to
the real line either by taking the absolute maximum or using a quadratic form -- with $p$-values
being computed from the analogous transformation of the normal distribution
\citep[see][for a hands-on introduction]{Hothorn+Hornik+VanDeWiel:2006}.

@@

Building on earlier work for the QUEST algorithm \citep{Loh+Shih:1997}, \citep{Loh:2002}
proposed the GUIDE algorithm blending trees with parametric regression models and
encompassing simpler classification and regression trees as special cases. Thus,
linear regression models could be fitted in the nodes of tree as well as
constant fits such as simple mean response $\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$,.
The tests for selecting the split variable are then based on the corresponding residuals,
e.g., for a simple linear regression \citep[as used in][]{Loh:2002}:
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In subsequent work other models together with an appropriate choice of residuals
have been applied such as in
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, 
%(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
quantile regression models \citep{Chaudhuri+Loh:2002}, 
and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015}, among others.

To construct a statistical test two additional transformation are carried out:
(1)~The residuals $r(Y,X,\hat{\beta})$ are binarized at zero, yielding an indicator for positive vs.\ negative residuals.
(2)~All possible split variable $Z$ are categorized, i.e., unless $Z$ is already categorical
it is split at its quartiles into four bins.
Subsequently, a $\chi^2$~test of independence is performed for the binarized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest 
$p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
Note that the split point in numeric variables $Z$ is searched over all possible splits,
not just the four bins that were constructed for the $\chi^2$~test.

@@

In the unbiased recursive partitioning literature this postpruning approach is also usef frequently
\citep[e.g., in][]{Loh+Vanichsetakul:1988,Loh+Shih:1997,Kim+Loh:2001,Loh:2002} and the $p$-values
from the association tests are only employed for selecting the split variable on a unified scale.
However, \cite{Hothorn+Hornik+Zeileis:2006} proposed to also use these $p$-values for a 
so-called \emph{prepruning} strategy which stops growing the tree as soon as no significant association
can be found in a given subgroup. This approach is the default in CTree and also in
MOB. However, \cite{Zeileis+Hothorn+Hornik:2008} also pointed out that a natural strategy for
postpruning in model-based partitioning is to use information criteria such as AIC (Akaike information
criterion) or BIC (Bayes information criterion), following the ideas of \cite{Su+Wang+Fan:2004}.

@@

\section{Unifying framework}\label{sec:unify}
Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
of the type of fitted models, the type of applied tests and the employed pruning strategy
as listed in Table~\ref{tab:combinations}. These settings represent the algorithms in their
original form. However, in principle all other combinations of the listed features are possible 
as well. As explained in Section~\ref{sec:pruning} the focus of this paper is on the testing 
strategies (second column of Table~\ref{tab:combinations}). They are here discussed for the case 
of parametric models together with prepruning as the properties of this pruning strategy
are already determined by the testing strategy. Subsequently combinations
with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
From here on the names CTree, MOB, and GUIDE are used to refer to the algorithm-specific
testing strategies rather than the complete algorithms.

@@

Even though CTree, MOB, and GUIDE differ in the specific tests they apply (see second column of
Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
basic structure as explained in Section~\ref{sec:generic_alg}. 
Thus they can be presented together in a unifying framework for testing strategies
in model-based unbiased recursive partitioning. 
In this framework each specific testing strategy can be obtained by combining the provided 
building blocks accordingly. But apart from that, variations of the commonly known testing 
strategies can be obtained as this general toolbox allows for all possible combinations of the 
building blocks. Moreover, by exchanging only one of them its impact can be investigated in detail.
These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
for the standard setup of CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which 
gives a more detailed overview of the differences in the second column of Table~\ref{tab:combinations}.

\subsection{Building blocks of testing strategies}\label{sec:buildingblocks}
The three following building blocks all refer to the way how the given data is prepared/transformed
before employing a statistical test to select a split variable.
\begin{itemize}
\item Variations of the goodness-of-fit measure:\\
Different variations of the disaggregated, observation-wise goodness-of-fit measure of the 
model that is obtained for the response $Y$ and possible covariates $X$ can be 
considered. This can be residuals $r(Y,X,\hat{\beta})$ as in the GUIDE algorithm or 
scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, more generally, a 
transformation function $h$ as in the CTree algorithm including both of the foregoing cases.
By including all available scores rather than just residuals the goodness of fit of the 
model is investigated regarding each parameter instead of just the mean. 
%Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.

@@

\item Categorization of covariates:\\
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
and 3rd quartile are considered to bin each possible split variable into four categories. In that way,
depending on whether the scores are binarized or not, a $\chi^2$~test or 1-way ANOVA can be applied
where the quartiles are the only possible split points while otherwise maximally selected splits 
over $Z$ are considered.

>>>>

\subsection{Generic algorithm}
\label{sec:generic_alg}
The basic idea of building a regression tree model is to partition the data into smaller 
and more homogeneous subgroups based on a set of covariates. Various tree 
algorithms have been developed, following essentially the same general structure.
Starting at the root of the tree, pertaining to the full available data sample, the
algorithms proceed in the following steps:
%
\begin{enumerate}
  \item For the current sample a (possibly simple) model is fitted by optimizing
        some objective function (or loss function) that reflects the goodness of fit.
  \item Among all available covariates one is selected as the split variable,
        choosing the split point such that goodness of fit is maximized in the
        resulting subgroups.
  \item Steps 1 and 2 are repeated within each subgroup until some stopping
        criterion is attained.
\end{enumerate}
%
The term ``model'' is used here in a very broad sense and encompasses not
only least-squares or maximum-likelihood models but also simple constant
fits such as means or average proportions.
Therefore, depending on the type of model, the employed objective function can 
for example be the sum of deviations from a typical/average value, but it can 
also be based on a model for a response along with potential covariates.
For instance, different types of residuals (or the signs thereof) can be employed
\citep[see e.g.,][]{Loh:2002} as well as rank sums or logrank scores
\citep[as in][]{Hothorn+Hornik+VanDeWiel:2006}.

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and each possible transformed split variable $g(Z_j)$ ($j = 1, \dots, m$).
The only requirement for the function $h$ is to depend on $Y$ in a permutation-symmetric way but
this encompasses ranks, scores, indicator functions, etc.\ and can also be multidimensional.
For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can capture location and scale of $Y$ via
$h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fitted to the response $Y$ with regressor variable(s)~$X$,
then a model-based transformation $h(Y) = s(Y,X,\hat{\beta})$ can be used, e.g.,
the residuals in a linear model with regression coefficients $\beta$. Moreover,
$s$ can be the score function pertaining to the objective function (or loss function)
$\ell$:
$$
s(Y, X, \beta) = \frac{\partial \ell(Y, X, \beta)}{\partial \beta}
$$
The estimate of the model parameters $\hat{\beta}$ is obtained by optimizing
the sum of the objective function $\ell$, aggregated over all observations.
This framework includes many different M-type estimators as special cases, including maximum 
likelihood and ordinary least squares estimation. For a $k$-dimensional parameter
$\beta$ the score function evaluated for the $i$-th observation $s(y_i, x_i, \beta)$
is also a $k$-dimensional vector, i.e., the gradient contribution of the $i$-th observation.
Thus, the $n \times k$-matrix consisting of these scores or gradient contributions for all 
$i = 1, \dots, n$ observations is a natural candidate for the observation-wise goodness-of-fit
measure as described in the previous Section~\ref{sec:generic_alg}.

@@

To test for independence of $h(Y)$ and $g(Z)$ CTree calculates a linear association test statistic,
following the framework of \cite{Strasser+Weber:1999}. The corresponding conditional expectation and
covariance given all permutations of the response variable can be calculated and used to 
standardize the test statistic. This standardized statistic has an asymptotic normal distribution
which is in fact multivariate if either of the transformation $h(Y)$ and/or $g(Z)$ is
multivariate. The actual test is carried out by mapping this standardized statistic to
the real line either by taking the absolute maximum or using a quadratic form -- with $p$-values
being computed from the analogous transformation of the normal distribution
\citep[see][for a hands-on introduction]{Hothorn+Hornik+VanDeWiel:2006}.

@@

Building on earlier work for the QUEST algorithm \citep{Loh+Shih:1997}, \cite{Loh:2002}
proposed the GUIDE algorithm blending trees with parametric regression models and
encompassing simpler classification and regression trees as special cases. Thus,
linear regression models could be fitted in the nodes of a tree as well as
constant fits such as simple mean response $\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$.
The tests for selecting the split variable are then based on the corresponding residuals,
e.g., for a simple linear regression \citep[as used in][]{Loh:2002}:
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In subsequent work other models together with an appropriate choice of residuals
have been applied such as in
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, 
%(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
quantile regression models \citep{Chaudhuri+Loh:2002}, 
and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015}, among others.

To construct a statistical test two additional transformations are carried out:
(1)~The residuals $r(Y,X,\hat{\beta})$ are binarized at zero, yielding an indicator for positive vs.\ negative residuals.
(2)~Each possible split variable $Z$ is categorized, i.e., unless $Z$ is already categorical
it is split at its quartiles into four bins.
Subsequently, a $\chi^2$~test of independence is performed for the binarized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest 
$p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
Note that the split point in numeric variables $Z$ is searched over all possible splits,
not just the four bins that were constructed for the $\chi^2$~test.

@@

In the unbiased recursive partitioning literature this postpruning approach is also used frequently
\citep[e.g., in][]{Loh+Vanichsetakul:1988,Loh+Shih:1997,Kim+Loh:2001,Loh:2002} and the $p$-values
from the association tests are only employed for selecting the split variable on a unified scale.
However, \cite{Hothorn+Hornik+Zeileis:2006} proposed to also use these $p$-values for a 
so-called \emph{prepruning} strategy which stops growing the tree as soon as no significant association
can be found in a given subgroup. This approach is the default in CTree and also in
MOB. However, \cite{Zeileis+Hothorn+Hornik:2008} also pointed out that a natural strategy for
postpruning in model-based partitioning is to use information criteria such as AIC (Akaike information
criterion) or BIC (Bayes information criterion), following the ideas of \cite{Su+Wang+Fan:2004}.

@@

\section{Unifying framework}\label{sec:unify}
Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
of the type of fitted models, the type of applied tests and the employed pruning strategy
as listed in Table~\ref{tab:combinations}. These settings represent the algorithms in their
original form. However, in principle all other combinations of the listed features are possible 
as well. As explained in Section~\ref{sec:pruning} the focus of this paper is on the testing 
strategies (second column of Table~\ref{tab:combinations}). They are here discussed for the case 
of parametric models together with prepruning as the properties of this pruning strategy
are already determined by the testing strategy. Subsequently, combinations
with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
From here on the names CTree, MOB, and GUIDE are used to refer to the algorithm-specific
testing strategies rather than the complete algorithms.

@@

Even though CTree, MOB, and GUIDE differ in the specific tests they apply (see second column of
Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
basic structure as explained in Section~\ref{sec:generic_alg}. 
Thus, they can be presented together in a unifying framework for testing strategies
in model-based unbiased recursive partitioning. 
In this framework each specific testing strategy can be obtained by combining the provided 
building blocks accordingly. But apart from that, variations of the commonly known testing 
strategies can be obtained as this general toolbox allows for all possible combinations of the 
building blocks. Moreover, by exchanging only one of them its impact can be investigated in detail.
These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
for the standard setup of CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which 
gives a more detailed overview of the differences in the second column of Table~\ref{tab:combinations}.

\subsection{Building blocks of testing strategies}\label{sec:buildingblocks}
The three following building blocks all refer to the way how the given data is prepared/transformed
before employing a statistical test to select a split variable.
\begin{itemize}
\item Variations of the goodness-of-fit measure:\\
Different variations of the disaggregated, observation-wise goodness-of-fit measure of the 
model that is obtained for the response $Y$ and possible covariates $X$ can be 
considered. This can be residuals $r(Y,X,\hat{\beta})$ as in the GUIDE algorithm or 
scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, more generally, a 
transformation function $h(Y)$ as in the CTree algorithm including both of the foregoing cases.
By including all available scores rather than just residuals the goodness of fit of the 
model is investigated regarding each parameter instead of just the mean. 
%Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.

@@

\item Categorization of covariates:\\
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd, 
and 3rd quartile are considered to bin each possible split variable into four categories. In that way,
depending on whether the scores are binarized or not, a $\chi^2$~test or 1-way ANOVA can be applied
where the quartiles are the only possible split points while otherwise maximally selected splits 
over $Z$ are considered.
